Evaluating sentinel pipe racks for monitoring lead release and optimizing corrosion control

### 2023-08-21 Rmarkdown/paper.Rmd
chemrxiv resubmission

Orthophosphate can limit lead contamination of tap water, but its benefits are difficult to quantify since lead concentrations are so site-specific. Sentinel homes serviced by lead pipe are ideal for monitoring orthophosphate treatment, but best practices dictate removal of lead once identified. The best sentinel homes, then, are often short-lived. Here we explore an alternative: recovered lead pipe racks supplied with distributed drinking water at locations throughout a water system. We also propose a strategy for analyzing the data based on the generalized additive model, which approximates time series as sums of smooth functions. Geometric mean lead release from pipe racks exhibited a pronounced dose-response, falling by `r round(abs(pcent_decrease$y))`% after an increase from 1 to 2 mg PO~4~ L^-1^, and then climbing by `r round(abs(pcent_increase$y))`% after a decrease to 1.5 mg PO~4~ L^-1^. Data from nine sentinel homes were consistent with those from pipe racks: geometric mean lead at the high orthophosphate dose was `r d$d_50`% of that at the low dose. Our results demonstrate sentinel pipe racks as a viable alternative to at-the-tap sampling for non-regulatory corrosion control monitoring. They also provide a Bayesian framework for quantifying changes in lead release that can incorporate information from multiple sources.

@@

Updated regulations on lead in drinking water promise to expedite replacement of lead service lines in Canada and the USA. Even afterwards, though, a substantial legacy of lead plumbing---including lead:tin solder and brass---will have to be managed. This will require careful control of drinking water chemistry to limit lead solubility and maintain durable corrosion scale.

@@

We present data from three separate racks, located at three sites within the Halifax Regional Municipality, a medium-sized Canadian city. We used a robust hierarchical Bayesian generalized additive model with continuous-time autoregressive errors [@trueman_comparing_2022] to estimate the effect on lead release of a dose increase from 1 to 2 mg PO~4~ L^-1^. Then, we used this estimate as a prior probability for the same effect in nine sentinel homes. Finally, we quantified the orthophosphate dose response of a subset of the pipe racks at 1, 2, and then 1.5 mg PO~4~ L^-1^. Our results provide a Bayesian framework for analyzing corrosion control treatment data, especially when they are collected as time series and have multiple sources.

@@

```{r summary-wq}
options(knitr.kable.NA = "")
wq %>% 
  group_by(param) %>% 
  mutate(unit = unfill_vec(unit)) %>% 
  ungroup() %>% 
  transmute(
    Parameter = unfill_vec(param),
    Unit = str_replace(unit, "ug", "µg"),
    Zone = zone,
    Median = round(med, 1), 
    `Lower quartile` = round(lq, 1),
    `Upper quartile` = round(uq, 1)
  ) %>% 
  knitr::kable(caption = "**Table 1.** Summary of water quality in pipe rack effluent by zone, with the exception of hardness, which was measured in distributed water samples.")
```

@@

Of the nine sentinel homes, seven were supplied by partial lead service lines (private lead, public copper) and the remaining two by copper service lines; all were located in Zone 2. At each sampling round, volunteer residents collected four consecutive 1L samples, starting with the first-draw after a minimum six-hour stagnation period. This 4 $\times$ 1L profile was followed first by a 10-minute flush of the plumbing and then by collection of a final 1L sample. All samples were collected in wide-mouth plastic bottles and acidified with trace metal-grade nitric acid. Volunteers were instructed to open their faucets all the way and not to adjust the flow rate while sampling (Figure S3).

@@

Even without an informative prior on $\beta$ (equation 2), however, the point-of-use data yielded a more precise estimate of the effect of doubling the orthophosphate dose, with a 95% credible interval that was approximately half as wide as the one estimated from the pipe racks. This is mainly a result of the differences in materials and lead levels; since the pipe racks represent lead release from lead pipe galvanically-coupled at both ends to brass, we expect high variability [@masters_inherent_2016]. And since the point-of-use data more probably represent lead release from a variety of sources---and at a much lower level---we expect lower variability [@masters_inherent_2016].

@@

A little more than a year after the orthophosphate dose was increased in Zone 1, it was decreased from 2 to 1.5 mg PO~4~ L^-1^ (Figure 3b). We used the sentinel pipe racks to determine the orthophosphate dose response in this zone, by estimating the effect of an increase from 1 to 2 mg PO~4~ L^-1^ and of a subsequent decrease to 1.5 mg PO~4~ L^-1^. But since the final decrease occurred in the spring---as water temperatures were increasing rapidly (Figure 3e)---we estimated the seasonal variation in lead release and added it as a separate term in the model to control for temperature effects [@masters_seasonal_2016]. Seasonality was more complex in Zone 2, perhaps due to the inverse seasonal pattern in aluminum (especially before the change in coagulation pH [@trueman_seasonal_2022]). And because the dose increase effect was estimated year-over-year in the full model (encompassing both zones) controlling for seasonality was less important.

@@

This result has implications for passivation-maintenance orthophosphate dosing strategies---that is, initiating treatment at a high orthophosphate dose to promote lead phosphate scale formation and then decreasing the dose once scale evolution has slowed [@schock1998lead]. Although lead solubility is predicted to increase with a decrease in orthophosphate, the effect on particulate lead is unclear: an established lead-phosphate scale, for instance, may be no less durable after a decrease in the orthophosphate dose. Passivation/maintenance dosing has the potential to conserve phosphorus, but it should be evaluated carefully to avoid unwanted increases in lead release at the maintenance dose or excess particulate lead at an unnecessarily high passivation dose [@zhao_formation_2018; @trueman_comparing_2022]. Here though, the dose response of lead release to orthophosphate was qualitatively similar to that predicted by solubility: lead release decreased when orthophosphate was increased and increased when orthophosphate was decreased.

@@

Point-of-use sampling is necessary to accurately quantify lead release into drinking water. But lead service line replacement, incomplete participation by residents in sampling programs, and changes to premises plumbing make it difficult to monitor the effectiveness of corrosion control over time this way. And while no simple apparatus can reliable quantify human exposure to lead, sentinel pipe racks offer an alternative to point-of-use sampling for non-regulatory monitoring. Especially when installed at multiple locations across a water distribution network, sentinel pipe racks can be used to understand how both anticipated and unexpected changes in water quality impact lead concentrations. We used them here to estimate the effect on lead release of changes in orthophosphate dose and coagulation process. By partitioning the variation in lead concentrations hierarchically---estimating global and location-level trends---we were better able to control for seasonality or other potential confounders before quantifying the effects of interest.

>>>>

Orthophosphate can limit lead contamination of tap water, but its benefits are difficult to quantify since lead concentrations are so site-specific. Sentinel homes serviced by lead pipe are ideal for monitoring orthophosphate treatment, but best practices dictate the removal of lead once identified. The best sentinel homes, then, are often short-lived. Here we explore an alternative: recovered lead pipe racks supplied with distributed drinking water at locations throughout a water system. We also propose a strategy for analyzing the data based on the generalized additive model, which approximates time series as sums of smooth functions. Geometric mean lead release from pipe racks exhibited a pronounced dose-response, falling by `r round(abs(pcent_decrease$y))`% after an increase from 1 to 2 mg PO~4~ L^-1^, and then climbing by `r round(abs(pcent_increase$y))`% after a decrease to 1.5 mg PO~4~ L^-1^. Data from nine sentinel homes were consistent with those from pipe racks: geometric mean lead at the high orthophosphate dose was `r d$d_50`% of that at the low dose. Our results demonstrate sentinel pipe racks as a viable alternative to at-the-tap sampling for non-regulatory corrosion control monitoring. They also provide a Bayesian framework for quantifying changes in lead release that can incorporate information from multiple sources.

@@

Updated regulations on lead in drinking water promise to expedite replacement of lead service lines in Canada and the USA. Even afterward, though, a substantial legacy of lead plumbing---including lead:tin solder and brass---will have to be managed. This will require careful control of drinking water chemistry to limit lead solubility and maintain durable corrosion scale.

@@

We present data from three separate racks, located at three sites within the Halifax Regional Municipality, a medium-sized Canadian city. We used a robust hierarchical Bayesian generalized additive model with continuous-time autoregressive errors [@trueman_comparing_2022] to estimate the effect on lead release of a dose increase from 1 to 2 mg PO~4~ L^-1^. Then, we used this estimate as a prior probability for the same effect in nine sentinel homes. Finally, we quantified the orthophosphate dose-response of a subset of the pipe racks at 1, 2, and then 1.5 mg PO~4~ L^-1^. Our results provide a Bayesian framework for analyzing corrosion control treatment data, especially when they are collected as time series and have multiple sources.

@@

```{r summary-wq}
options(knitr.kable.NA = "")
wq %>% 
  group_by(param) %>% 
  mutate(unit = unfill_vec(unit)) %>% 
  ungroup() %>% 
  transmute(
    Parameter = unfill_vec(param),
    Unit = str_replace(unit, "ug", "µg"),
    Zone = zone,
    Median = round(med, 1), 
    `Lower quartile` = round(lq, 1),
    `Upper quartile` = round(uq, 1)
  ) %>% 
  knitr::kable(caption = "**Table 1.** Summary of water quality in pipe rack effluent by zone, except for hardness, which was measured in distributed water samples.")
```

@@

Of the nine sentinel homes, seven were supplied by partial lead service lines (private lead, public copper) and the remaining two by copper service lines; all were located in Zone 2. At each sampling round, volunteer residents collected four consecutive 1L samples, starting with the first-draw after a minimum six-hour stagnation period. This 4 $\times$ 1L profile was followed first by a 10-minute flush of the plumbing and then by collection of a final 1L sample. All samples were collected in wide-mouth plastic bottles and acidified with trace metal-grade nitric acid. Volunteers were instructed to open their faucets all the way and not to adjust the flow rate while sampling, but they were not instructed to measure flow rate (Figure S3).

@@

Even without an informative prior on $\beta$ (equation 2), however, the point-of-use data yielded a more precise estimate of the effect of doubling the orthophosphate dose, with a 95% credible interval that was approximately half as wide as the one estimated from the pipe racks. This is mainly a result of the differences in materials and lead levels; since the pipe racks represent lead release from lead pipe galvanically coupled at both ends to brass, we expect high variability [@masters_inherent_2016]. And since the point-of-use data more probably represent lead release from a variety of sources---and at a much lower level---we expect lower variability [@masters_inherent_2016].

@@

A little more than a year after the orthophosphate dose was increased in Zone 1, it was decreased from 2 to 1.5 mg PO~4~ L^-1^ (Figure 3b). We used the sentinel pipe racks to determine the orthophosphate dose-response in this zone, by estimating the effect of an increase from 1 to 2 mg PO~4~ L^-1^ and of a subsequent decrease to 1.5 mg PO~4~ L^-1^. But since the final decrease occurred in the spring---as water temperatures were increasing rapidly (Figure 3e)---we estimated the seasonal variation in lead release and added it as a separate term in the model to control for temperature effects [@masters_seasonal_2016]. Seasonality was more complex in Zone 2, perhaps due to the inverse seasonal pattern in aluminum (especially before the change in coagulation pH [@trueman_seasonal_2022]). And because the dose increase effect was estimated year-over-year in the full model (encompassing both zones) controlling for seasonality was less important.

@@

This result has implications for passivation-maintenance orthophosphate dosing strategies---that is, initiating treatment at a high orthophosphate dose to promote lead phosphate scale formation and then decreasing the dose once scale evolution has slowed [@schock1998lead]. Although lead solubility is predicted to increase with a decrease in orthophosphate, the effect on particulate lead is unclear: an established lead-phosphate scale, for instance, may be no less durable after a decrease in the orthophosphate dose. Passivation/maintenance dosing has the potential to conserve phosphorus, but it should be evaluated carefully to avoid unwanted increases in lead release at the maintenance dose or excess particulate lead at an unnecessarily high passivation dose [@zhao_formation_2018; @trueman_comparing_2022]. Here though, the dose-response of lead release to orthophosphate was qualitatively similar to that predicted by solubility: lead release decreased when orthophosphate was increased and increased when orthophosphate was decreased.

@@

Point-of-use sampling is necessary to accurately quantify lead release into drinking water. But lead service line replacement, incomplete participation by residents in sampling programs, and changes to premises plumbing make it difficult to monitor the effectiveness of corrosion control over time this way. And while no simple apparatus can reliably quantify human exposure to lead, sentinel pipe racks offer an alternative to point-of-use sampling for non-regulatory monitoring. Especially when installed at multiple locations across a water distribution network, sentinel pipe racks can be used to understand how both anticipated and unexpected changes in water quality impact lead concentrations. We used them here to estimate the effect on lead release of changes in orthophosphate dose and coagulation process. By partitioning the variation in lead concentrations hierarchically---estimating global and location-level trends---we were better able to control for seasonality or other potential confounders before quantifying the effects of interest.
